Relevance-based Quantization of Scattering Features for Unsupervised Mining of Environmental Audio

### 2018-08-17 paper/revision/responseToReviewers.tex

\item \emph{I suggest that they only refer to ecoacoustics in the discussion by mentionning that this could be tested on other audio scenes as those recorded in natural environments.}

$\rightarrow$ The abstract introduction, background section and conclusion have been rewritten to widen the scope of application of the proposed method and suggest the ecoacoustic as potential area of interest for further research.

\item \emph{The dataset is too quickly described. I had to read ref\#36 to understand how it was built and made of. It would be nice to have an example of a labelled recording either as a spectrogram or scalogram to estimate the complexity of the scenes recorded. The event classes should also be detailed.}

$\rightarrow$ We have listed the classes from the DCASE taxonomy in full in Section 6.1. In addition, we have included the scalogram of a labeled recording (\emph{park04}} as Figure 3.

@@

\item \emph{The authors present their technique as opposed to the BOF approach, that is considered "state of the art". BOF maybe a popular technique but the fact that gets rid of the temporal structure makes it useless in many tasks. In fact, in DCASE 2013 BOF was the chosen baseline system but no participant proposed that technique. Certainly, BOF allows the authors to make stronger the point about describing an acoustic scene from a few distinct events, but perhaps they should avoid to claim BOF as "the" state of the art in this problem.}

$\rightarrow$
As discussed above in the response to the first reviewer, the application in the DCASE 2013 paper, classification, is different from the one proposed in our work, similarity retrieval. For classification, the BOF approach is indeed not the state of the art. However, in acoustic scene similarity retrieval, we are not aware of any systems which outperform the BOF. As such, it is indeed the state of the art for the similarity retrieval task. The manuscript has been revised to clarify this point.

\item \emph{It would be interesting to see the contribution to the results of the Gammatone wavelet in comparison to Morlet or another more classical wavelet.}

$\rightarrow$ VL $\checkmark$ We have not re-run experiments with Morlet wavelets instead of Gammatone wavelets, for two reasons. First, Morlet wavelets do not fit the whole narrative of the paper as well as Gammatone wavelets: while the arguments in favor of Morlet wavelets arise from harmonic analysis (Heisenberg time-frequency uncertainty tradeoff), those in favor of Gammatone wavelets arise from auditory neurophysiology. Because we work in the context of computational auditory scene analysis, \emph{a fortiori} in an unsupervised setting, we believe that it is preferable to borrow from a well-studied auditory model (Gammatone) rather than from a general-purpose operator (Morlet). The second reason why we did not conduct a systematic benchmark between Morlet and Gammatone is that, in previous experiments involving supervised classification of musical instruments with scattering coefficients, Gammatone wavelets appeared to outperform Morlet wavelets by a small margin, \ie{} from $80\%$ and $79\%$ test accuracy respectively.
In all likelihood, the effect of the specific choice of wavelet is also small in the case of unsupervised similarity retrieval between auditory scenes, and possibly not statistically significant given the limited sample size of the DCASE 2013 dataset. We have expanded Section 3.4 to highlight these considerations. In addition, we have included a mathematical description of the Gammatone wavelet filter bank employed in this study as supplementary material. Indeed, the design of our Gammatone wavelet is slightly different than in the reference papers of Venkitaraman and Adiga, because, unlike them, we need to account for the influence of both the quality factor $Q$ and the polynomial degree parameter $N$. It is our hope that the proof of the formula linking $Q$ and $N$ to the parameter $\alpha$ in the Gammatone function will be beneficial to the signal processing community at large.

### 2018-09-12 paper/proof/13636_2018_138_MOESM1.tex
sup material

\documentclass[smallextended]{svjour3}
%\usepackage{cite}
% *** GRAPHICS RELATED PACKAGES ***
%
  \usepackage[pdftex]{graphicx}
%
\usepackage[cmex10]{amsmath}

@@

\usepackage[backend=bibtex,style=numeric-comp,sorting=none,firstinits=true]{biblatex}
\bibliography{biblio}
%\renewcommand{\baselinestretch}{2}

@@

\begin{document}
%
% paper title
% can use linebreaks \\ within to get better formatting as desired
% Do not put math or special symbols in the title.
%\title{An evaluation framework for event detection using a morphological model of acoustic scenes}
\title{Relevance-based Quantization of Scattering Features for Unsupervised Mining of Environmental Audio}
\titlerunning{Relevance-based Quantization of Scattering Features}

\author{Vincent Lostanlen         \and
		Gr\'egoire Lafay  \and Joakim And\'en  \and Mathieu Lagrange.}

\institute{Vincent Lostanlen \at
              \email{vincent.lostanlen@nyu.edu}
\and
			Gr\'egoire Lafay \at
              \email{lafaygregoire@gmail.com}
\and
			Joakim And\'en \at
              \email{janden@flatironinstitute.org}           \and
			Mathieu Lagrange \at
              \email{mathieu.lagrange@cnrs.fr}
}

\date{Received: date / Accepted: date}
% The correct dates will be entered by the editor

% make the title area
\maketitle

% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\begin{abstract}
The emerging field of computational acoustic monitoring aims at retrieving high-level information from acoustic scenes recorded by some network of sensors. These networks gather large amounts of data requiring analysis. To decide which parts to inspect further, we need tools that automatically mine the data, identifying recurring patterns and isolated events. This requires a similarity measure for acoustic scenes that does not impose strong assumptions on the data.

The state of the art in audio similarity measurement is the ``bag-of-frames'' approach, which models a recording using summary statistics of short-term audio descriptors, such as mel-frequency cepstral coefficients (MFCCs). They successfully characterise static scenes with little variability in auditory content, but cannot accurately capture scenes with a few salient events superimposed over static background.
To overcome this issue, we propose a two-scale representation which describes a recording using clusters of scattering coefficients. The scattering coefficients capture short-scale structure, while the cluster model captures longer time scales, allowing for more accurate characterization of sparse events. Evaluation within the acoustic scene similarity framework demonstrates the interest of the proposed approach.
\keywords{
unsupervised learning \and data mining \and acoustic signal processing \and wavelet transforms \and audio databases \and content-based retrieval \and nearest neighbor searches \and acoustic sensors \and environmental sensors.}
% \PACS{PACS code1 \and PACS code2 \and more}
\end{abstract}

\section{Introduction}

The amount of audio data recorded from our sonic environment has grown considerably over the past decades.
In order to measure the effect of human activity and climate change on animal biodiversity~\cite{sueur2015ecoacoustics}, researchers have recently undertaken a massive deployment of acoustic sensors throughout the world~\cite{warren2006urban, NessSST13, stowell13b}.
In addition, recent work has explored acoustic monitoring for characterization of human pleasantness in urban areas~\cite{guyot2005urban, ricciardi2015sound}, as well as the prediction of annoyance due to traffic~\cite{gloaguen}.
Since they bear a strong societal impact and raise many scientific challenges, we believe that these applications are of considerable interest for signal processing community.

%These questions are explored in an emerging field of computational bioacoustics, known as ecoacoustics, which is the interdisciplinary study of relationships between natural or anthropogenic sounds and their environments at the intersection of ecology, acoustics and computer science . %~\cite{wimmer2013sampling}

An important problem is that manually analysing the recorded data to identify the quantities of interest is very costly. Some sort of pre-screening is therefore required to reduce the need for human expert listening and annotation.
To this aim, the most straightforward approach is to specify a closed set of sound classes, such as sounds classes expected to appear near the acoustic sensors. Computational models are then trained  for these classes which are used to automatically annotate recordings~\cite{quteprints96243}. A given time interval (\eg{} a single day) is then represented by the number of events detected during that interval for each class. This allows the scientist to drastically reduce the amount of information requiring manual processing. However, this approach has two drawbacks. First, it relies on trained models whose prediction on unseen data -- \ie{} sensors outside of the training set -- is prone to errors. Secondly, and more importantly, it is based on prior knowledge and thus cannot be considered for exploratory analysis, in which quantities of interest have yet to be defined.

To identify which parts need human inspection, one needs tools that are able to detect both recurring patterns and sparsely distributed events. Identifying recurring patterns allows the user to focus on certain time points for manual annotation, while detection of more rare structures enables discovery of unforeseen phenomena.

With this aim, we need to design an algorithm for acoustic similarity retrieval, where the audio fragments judged ``most similar'' to a given query recording must be extracted from some larger dataset. To construct such an algorithm, we are required to represent an audio recording in a way that captures its distinctive qualities. A widespread choice of representation is the bag of frames~\cite{aucouturier2007bag}, which describes an auditory scene recording using summary statistics of short-time features. Unfortunately, the bag-of-frames approach only captures the average structure of the scene, so the approach often fails when presented with highly dynamic scenes or those characterised by a few distinct sound events sparsely distributed over time. Furthermore, experiments in cognitive psychology~\cite{dubois2006cognitive} and cognitive neuroscience~\cite{nelken2004processing} suggest that human acoustic perception is highly sensitive to such isolated sound events. We believe that the failure to model such distinct events is one of the reasons why the bag-of-frames representation is insufficient~\cite{lagrange:hal-01082501}.

Solving the acoustic similarity retrieval first requires the ability to capture meaningful signal structure at small time scales.
This is often achieved using mel-frequency cepstral coefficients (MFCCs).
Originally developed for speech processing~\cite{davis-mermelstein}, MFCCs have recently found wider use in music information retrieval~\cite{logan} and environmental audio processing~\cite{aucouturier2007bag}.
A richer representation, the scattering transform, has enjoyed significant success in various audio~\cite{Anden2014} and biomedical~\cite{chudacek} signal classification tasks.
Its structure is that of a convolutional neural network~\cite{lee,lostanlen-deep-spiral,soundnet,arandjelovic-zisserman}, but with fixed filters.
Specifically, it alternates convolutions with wavelet filters and pointwise nonlinearities to ensure time-shift invariance and time-warping stability~\cite{Mallat2012}.

For our task, one advantage of the scattering transform is that it does not require a training step, allowing for a wider range of applications compared to learned features. Indeed, for data mining of previously unheard datasets, the properties of relevant audio structures remains to be defined, leading to an unsupervised setting.

In this work, we propose a new model for acoustic scenes, where the signal is represented at sub-second scales by scattering transforms, while larger scales are captured by a cluster model. This unsupervised model quantizes the scattering coefficients into a given number of clusters. These clusters are then used to define a set of distances for acoustic similarity retrieval. Evaluating this approach on a scene retrieval task, we obtain significant improvements over traditional bag-of-frames and summary statistics models applied both to MFCCs and scattering coefficients.

Motivations of the proposed approach and a brief review of the state of the art in acoustic scene modeling are given in Section~\ref{sec:soa}. We describe the scattering transform in Section~\ref{sec:scattering}, discuss feature post-processing in Section~\ref{sec:design} and propose a cluster-based scene description in Section~\ref{sec:object}. Section~\ref{sec:experiments} describes several experiments for the acoustic scene similarity retrieval task. Results are reported in Section~\ref{sec:results}.

\section{Background} \label{sec:soa}

%Computational bioacoustics refers to the numerical investigation of sound production, dispersion and reception in animals, including humans. A recent paradigm processes the audio stream in a holistic way over large time scales, without assuming that a single species is present throughout the recording. Automated systems within this paradigm attempt to infer global properties of bioacoustic scenes, including biodiversity indices~\cite{Bardeli2010} and migration patterns~\cite{Obrist2010}. They also mark time intervals of particular interest for detailed human inspection~\cite{rosenstock2002landbird}.

Caracterization of the similarity between audio recordings be they at the scale of the minute, the hour, the day or larger is of interest for many applications areas involving acoustic monitoring such as urban sound environment analysis and ecoacoustics. In this context, a classical approach is the bag of frames (BoF), first applied to the problem by Aucouturier et al.~\cite{aucouturier2007bag}.
It models an auditory scene using high-level summary statistics computed from local features, typically implemented by Gaussian mixture models (GMMs) of MFCCs.

It is worth mentionning that this task typically falls into a unsupervised paradigm where no prior knowldge is used to model a given scene. For each scene $s$, a model is compute $M_s$ is computed. The similarity among the scene $s_1$ and $s_2$ is computed as the similarity, see Section~\ref{sec:experiments} for further details. The BoF approach is also widely used in a supervised fashion for solving a classification task~\cite{7100934}. In this case, each class of scenes from a given typology, say \{\emph{park, boulevard, square}\}, is modeled by a GMM trained on scenes taken from a training set. In order to predict the class of a given scene $s$, the likelihood each model given the scene are computed. The scene is then labeled park if the likelihood of the GMM trained on park scenes is higher than all other likelihoods.

While BoF has largely been superseded by more sophisticated methods for the task of acoustic scene classification~\cite{7100934}, it remains the best-performing model for acoustic scene similarity retrieval.
Though, this representation was recently shown to perform comparably to direct averaging of the features for a variety of datasets~\cite{lagrange:hal-01082501}.
This contrasts with the typical morphology of acoustic scenes, a ``skeleton of events on a bed of textures,'' where a few discrete sound events are superimposed upon a stationary acoustic background~\cite{nelken2013}.
Such events are not well-characterised by summarizing short-term features, but are better described by large-scale temporal evolution of auditory scenes.
The latter approach should therefore prove more fruitful in measuring auditory scene similarity.

This statement has some support in auditory psychology as well as sound synthesis based on summary statistics~\cite{mcdermott2013summary}. Studies in the cognitive psychology of urban sound environments have shown that global sound level (perceived or measured) is not sufficient to fully characterise an acoustic scene~\cite{guyot2005urban,kang2006urban}. Instead, cognitive processes such as sound environment quality perception~\cite{dubois2006cognitive} or loudness judgment~\cite{kuwano_memory_2003} seem to rely upon higher-level cognitive attributes. These typically include the identities of the sound sources which constitute the scene. It has been shown that, if available, the complete description of the scene in terms of event occurrences is powerful enough to reliably predict high-level cognitive classes. For example, in urban areas the presence of birds is likely to be heard in parks and are therefore strong pleasantness indicators. Consequently, research in sound perception is now strongly focused on the contribution of specific sound sources in the assessment of sound environments~\cite{ricciardi2015sound,lavandier2006contribution}. Although the complete set of events occurring within a given auditory stream may not be discernable even to human expects, research has shown that a small set of events (so-called markers) suffice to reliably predict many high-level attributes.

From a cognitive psychology perspective, the consensus is therefore that only a few distinct events are sufficient to describe an auditory scene, in contrast to BoF models which treat each observation separately and do not capture their temporal structure. A method that takes this knowledge into account could therefore have potential for great impact in acoustic scene modeling, given a rich enough representation of these distinct events.

\section{Wavelet scattering \label{sec:scattering}}

Local invariance to time-shifting and stability to time-warping are necessary when representing acoustic scenes for similarity measurement.
The scattering transform is designed to satisfy these properties while retaining high discriminative power.
It is computed by applying auditory and modulation wavelet filter banks alternated with complex modulus nonlinearities.

\subsection{Invariance and stability in audio signals}
The notion of invariance to time-shifting plays an essential role in acoustic scene similarity retrieval.
Indeed, recordings may be shifted locally in time without affecting similarity to other recordings.
To discard this superfluous source of variability, signals are first mapped into a time-shift invariant feature space. These features are then used to calculate similarities. Since the features ensure invariance, it does not have to be learned when constructing the desired similarity measure.

Formally, given a signal $\boldsymbol{x}(t)$, we would like its translation $\boldsymbol{x_c}(t) = \boldsymbol{x}(t-c)$ to be mapped to the same feature vector provided that $|c| \ll T$ for some maximum duration $T$ that specifies the extent of the time-shifting invariance. We can also define more complicated transformations by letting $c$ vary with $t$. In this case, we have $\boldsymbol{x_\tau}(t) = \boldsymbol{x}(t-\tau(t))$ for some function $\tau$, which performs a time-warping of $\boldsymbol{x}(t)$ to obtain $\boldsymbol{x_\tau}(t)$. Time-warpings model various changes, such as small variations in pitch, reverberation, and rhythmic organization of events. These make up an important part of intra-class variability among natural sounds, so representations must be robust with respect to such transformations.

The wavelet scattering transform, described below, has both of these desired properties: invariance to time-shifting and stability to time-warping. The stability condition can be formulated as a Lipschitz continuity property, which guarantees that the feature transforms of $\boldsymbol{x}(t)$ and $\boldsymbol{x_\tau}(t)$ are close together if $|\tau'(t)|$ is bounded by a small constant \cite{Mallat2012}.

\subsection{Wavelet scalogram}
Our convention for the Fourier transform of a continuous-time signal $\boldsymbol{x}(t)$ is $\boldsymbol{\hat{x}}(\omega) = \int_{-\infty}^{+\infty} x(t) \exp(- \mathrm{i} 2\pi \omega t) \, \mathrm{d}t$.
Let $\boldsymbol{\psi}(t)$ a complex-valued analytic bandpass filter of
central frequency $\xi_1$ and bandwidth $\xi_1/Q_1$, where $Q_1$ is the quality factor of the filter.
A filter bank of wavelets is built by dilating $\boldsymbol{\psi}(t)$
according to a geometric sequence of scales $2^{\gamma_1/Q_1}$, obtaining
\begin{equation}
\boldsymbol{\psi_{\gamma_1}}(t) = 2^{-\gamma_1/Q_1} \boldsymbol{\psi}(2^{-\gamma_1/Q_1} t)\mbox{.}
\end{equation}
The variable $\gamma_1$ is a scale (an inverse log-frequency) taking integer values between $0$ and $(J_1 Q_1 - 1)$, where $J_1$ is the number of octaves spanned by the filter bank.
For each $\gamma_1$, the wavelet $\boldsymbol{\psi_{\gamma_1}}(t)$
has a central frequency of $2^{-\gamma_1/Q_1}\xi_1$ and a bandwidth of $2^{-\gamma_1/Q_1}\xi_1/Q_1$ resulting in the same quality factor $Q_1$ as $\boldsymbol{\psi}$.
In the following, we set $\xi_1$ to $20~\mathrm{kHz}$, $J_1$ to $10$, and the quality factor $Q_1$, which is also the number of wavelets per octave, to $8$. This results in the wavelet filters covering the whole range of human hearing, from $20~\mathrm{Hz}$ to $20~\mathrm{kHz}$. Setting $Q_1 = 8$ results in filters whose bandwidth approximates an equivalent rectangular bandwidth (ERB) scale~\cite{Fastl2007}.

The wavelet transform of an audio signal
$\boldsymbol{x}(t)$ is obtained by convolution with all wavelet filters.
Applying a pointwise complex modulus the transform yields
the wavelet scalogram
\begin{equation}
\boldsymbol{x_1}(t, \gamma_1)
= \vert \boldsymbol{x} \ast \boldsymbol{\psi_{\gamma_1}} \vert (t)\mbox{.}
\end{equation}
The scalogram bears resemblance to the constant-Q transform (CQT),
which is derived from the short-term Fourier transform (STFT) by averaging the frequency
axis into constant-Q subbands of central frequencies $2^{-\gamma_1/Q_1}\xi_1$.
Indeed, both time-frequency representations are indexed by time $t$ and log-frequency $\gamma_1$.
However, contrary to the CQT, the scalogram reaches a better time-frequency localization across the whole
frequency range, whereas the temporal resolution of the traditional CQT is fixed by the support of the STFT analyzing window. %\cite{Brown1992}
Therefore, the scalogram has a better temporal localization at high
frequencies than the CQT, at the expense of a greater computational cost
since the inverse fast Fourier transform routine must be called for each wavelet $\boldsymbol{\psi_{\gamma_1}}$ in the filter bank.
However, this allows us to observe amplitude modulations at fine temporal scales in the scalogram, down to $2Q_1/\xi_1$ for $\gamma_1 = 0$, of the order of $1\,\textrm{ms}$ given the aforementioned values of $Q_1$ and $\xi_1$.

To obtain the desired invariance and stability properties, the scalogram is averaged in time using a lowpass filter $\boldsymbol{\phi}(t)$ with cut-off frequency $1/T$ (and approximate duration $T$), to get
\begin{equation}
\mathbf{S_1}\boldsymbol{x}(t, \gamma_1) = \boldsymbol{x_1}(\cdot, \gamma_1) \ast \boldsymbol{\phi}(t),
\end{equation}
which is known as the set of first-order scattering coefficients. They capture the average spectral envelope of $\boldsymbol{x}(t)$ over scales of duration $T$ and where the spectral resolution varying with constant Q. In this way, they are closely related to the mel-frequency spectrogram and related features, such as MFCCs.

\subsection{Extracting modulations with second-order scattering}
In auditory scenes, short-time amplitude modulations may be caused by a variety of rapid mechanical interactions, including collision, friction, turbulent flow, and so on.
At longer time-scales, they also account for higher-level attributes of sound, such as prosody in speech or rhythm in music.
Although they are discarded while filtering $\boldsymbol{x_1}(t,\gamma_1)$ into the time-shift invariant representation $\mathbf{S_1}\boldsymbol{x}(t,\gamma_1)$, they can be recovered from $\boldsymbol{x_1}(t,\gamma_1)$ by a second wavelet transform and another complex modulus.

We define second-order wavelets $\boldsymbol{\psi_{\gamma_2}}(t)$ in the same way as the first-order wavelets, but with parameters $\xi_2$, $J_2$, and $Q_2$. Consequently, they have central frequencies $2^{-\gamma_2/Q_2}\xi_2$ for $\gamma_2$ taking values between $0$ and $(J_2 Q_2 - 1)$. While this abuses notation slightly, the identity of the wavelets should be clear from context.
The amplitude modulation spectrum resulting from a wavelet modulus decomposition using these second-order wavelets is then
\begin{equation}
\boldsymbol{x_2}(t,\gamma_1,\gamma_2) =
\vert \boldsymbol{x_1} \ast \boldsymbol{\psi_{\gamma_2}} \vert(t,\gamma_1).
\end{equation}
In the following, we set $\xi_2$ to $2.5\,\mathrm{kHz}$, $Q_2$ to $1$, and $J_2$ to $12$. Lastly, the low-pass filter $\boldsymbol{\phi}(t)$ is applied to $\boldsymbol{x_2}(t, \gamma_1, \gamma_2)$ to guarantee local invariance to time-shifting, which yields the second-order scattering coefficients
\begin{equation}
\mathbf{S_2}\boldsymbol{x}(t,\gamma_1,\gamma_2) =
\boldsymbol{x_2}(\cdot,\gamma_1,\gamma_2) \ast \boldsymbol{\phi}(t).
\end{equation}

The scattering transform $\mathbf{S}\boldsymbol{x}(t,\gamma)$ consists of the concatenation of first-order coefficients $\mathbf{S_1}\boldsymbol{x}(t,\gamma_1)$ and second-order coefficients $\mathbf{S_2}\boldsymbol{x}(t,\gamma_1,\gamma_2)$ into a feature matrix $\mathbf{S}\boldsymbol{x}(t,\gamma)$, where $\gamma$ denotes either $\gamma_1$ or $(\gamma_1,\gamma_2)$. While higher-order scattering coefficients can be calculated, for the purposes of our current work, the first and second order are sufficient.
Indeed, higher-order scattering coefficients have been shown to contain reduced energy and are therefore of limited use~\cite{irene}.

\subsection{Gammatone wavelets}
\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{gammatones}
\caption{
\label{fig:gammatones}
Gammatone wavelets $\boldsymbol{\psi}(t)$ in the time domain with quality factors (a) $Q = 4$ and (b) $Q = 1$.
Oscillations (red, blue) are the real and imaginary parts. The envelope (yellow) is the complex modulus.}
\end{center}
\end{figure}
Wavelets
$\boldsymbol{\psi_{\gamma_1}}(t)$ and $\boldsymbol{\psi_{\gamma_2}}(t)$ are designed as fourth-order Gammatone
wavelets with one vanishing moment~\cite{Venkitaraman2014}, and are shown in Figure~\ref{fig:gammatones}.
In the context of auditory scene analysis, the asymmetric envelopes of Gammatone wavelets are more biologically plausible than the symmetric, Gaussian envelopes of the more widely used Morlet wavelets.
Indeed, it allows to reproduce two important psychoacoustic effects in the mammalian cochlea: the asymmetry of temporal masking and the asymmetry of spectral masking~\cite{Fastl2007}.
The asymmetry of temporal masking is the fact that a masking noise has to be louder if placed after the onset of a stimulus rather than before.
Likewise, because critical bands are skewed towards higher frequencies, a masking tone has to be louder if it is above the stimulus in frequency rather than below.
It should also be noted that Gammatone wavelets follow the typical amplitude profile of natural sounds, beginning with a relatively sharp attack and ending with a slower decay.
As such, they are similar to filters discovered automatically by unsupervised encoding of natural sounds~\cite{Smith2006,sainath2015interspeech}.
In addition, Gammatone wavelets have proven to outperform Morlet wavelets on a benchmark of supervised musical instrument classification from scattering coefficients \cite{lostanlen2017phd}.
This suggests that, despite being hand-crafted and not learned, Gammatone wavelets provide a sparser time-frequency representation of acoustic scenes compared to other variants.  More information can be found in Annex \ref{sec:annex}.

\section{Feature design}
\label{sec:design}

Before constructing models for similarity estimation, it is beneficial to process scattering coefficients to improve invariance, normality, and generalization power.
In this section, we review two transformations which achieve these properties: logarithmic compression and standardisation.

\subsection{Logarithmic compression}
\label{sec:logcomp}

Many algorithms in pattern recognition, including nearest neighbor classifiers and support vector machines, tend to work best when all features follow a standard normal distribution across all training instances~\cite{Hsu2003}.
Yet the distribution of the scattering coefficients is skewed towards larger values. We can reduce this skewness by applying a pointwise concave transformation to all coefficients. In particular, we find that the logarithm performs particularly well in this respect.
Figure~\ref{fig:histograms} shows the distribution of an arbitrarily chosen scattering coefficient over the DCASE 2013 dataset, before and after logarithmic compression.

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{compression}
\caption{
\label{fig:histograms}
Histogram of values taken by the first-order scattering coefficient $\mathbf{S}\boldsymbol{x}(\gamma)$, corresponding to a central acoustic frequency of $302\,\mathrm{Hz}$,
(a) before and (b) after logarithmic compression.}
\label{fig:compression}
\end{center}
\end{figure}

Taking the logarithm of a magnitude spectrum is ubiquitous in audio signal processing.
Indeed, it is corroborated by the Weber-Fechner law in psychoacoustics, which states that the sensation of loudness is roughly proportional to the logarithm of the acoustic pressure.
We must also recall that the measured amplitude of sound sources often decays polynomially with the distance to the microphone--a source of spurious variability in scene classification.
Logarithmic compression linearise this dependency, facilitating the construction of powerful invariants at the classifier stage.

\subsection{Standardisation}
\label{sec:stand}

Let $\mathbf{S}\boldsymbol{x}(\gamma,n)$ be a dataset, where $\gamma$ and $n$ denote feature and sample indices, respectively.
Many algorithms operate better on features which have zero mean and unit variance to avoid mismatch in numeric ranges~\cite{Hsu2003}.
To standardise $\mathbf{S}\boldsymbol{x}(\gamma,n)$, we subtract the sample mean vector $\mu[\mathbf{S}\boldsymbol{x}(\gamma)]$ from $\mathbf{S}\boldsymbol{x}(\gamma,n)$ and divide the result by the sample standard deviation vector $\sigma[\mathbf{S}\boldsymbol{x}] (\gamma)$.
The vectors $\mu[\mathbf{S}\boldsymbol{x}(\gamma)]$ and $\sigma[\mathbf{S}\boldsymbol{x}](\gamma)$ are estimated from the entire dataset.

\section{Acoustic scene similarity retrieval}
\label{sec:object}

%As part of the aforementioned research areas and applications, the emerging field of \emph{Acoustic Scene Analysis} (also called \emph{Sound Scene Analysis})~\cite{Stowell15} aims to develop approaches and systems for the automatic analysis of environmental sounds and soundscapes (originating both from urban or nature environments). While research methodologies in related fields such as Automatic Speech Recognition (ASR)~\cite{Rabiner93} and Music Information Retrieval (MIR)~\cite{Muller07} are now well established, research addressing Acoustic Scene Analysis remains relatively young.

%From a data processing point of view, a holistic scheme, such as the bag-of-frame approach~\cite{aucouturier2007bag}, has a simplicity, but clearly face poor performance on realistic conditions~\cite{lagrange:hal-01082501}.

As discussed in Section~\ref{sec:soa}, results in sound perception suggest the appropriateness of source-driven representations of auditory scenes for predicting high-level properties. While this can be addressed in the supervised case using late integration of discriminative classifiers~\cite{Anden2014}, this is not directly feasible in the unsupervised case. As the detection of events is still an open problem~\cite{7100934}, we consider in this paper a generic quantization scheme in order to identify and represent time intervals of the scene that are coherent, thus likely to be dominated by a given source of interest.

Given a set of $d$-dimensional feature vectors $X_u = \{x_1^u, \ldots, x_L^u\}$, extracted from the scene $s_u$, where $u=\lbrace 1,2,\ldots,U\rbrace$, we would like to partition $X_u$ into a set $C_u = \{c^u_1, \ldots, c^u_M\}$ of $M$ clusters. This partition is obtained by minimizing the variance of each cluster and known as a $k$-means clustering~\cite{lloyd}.
Each scene $s_u$ is then described by a set of clusters $C_u$. Note that this quantization approach differs from unsupervised learning schemes such as the ones studied in~\cite{bisot2016acoustic}, where the scene features are projected in a dictionary learned from the entire dataset. Here, with the aim of better balancing the influence of salient sound events and texture-like sounds on the final decision, the similarity between two scenes is computed based on the similarity of their centroids.

The similarity between the scene centroids $\mu_m^u$ over the entire dataset, is computed using a radial basis function (RBF) kernel $K$ combined with a local scaling method~\cite{selfTuneManor2004}:
\begin{equation}
\label{eq:kc}
K_{mn}^{uv} = \exp\left( - \dfrac{\Vert \mu_m^u - \mu_n^v \Vert^2}{\Vert \mu_m^u - \mu_{m,q}^u \Vert \Vert \mu_n^v - \mu_{n,q}^v \Vert} \right).
\end{equation}
Here, $\mu_{m,q}^u$ and $\mu_{n,q}^v$ are the $q^{\textrm{th}}$ nearest neighbors to the centroids $\mu_m^u$ and $\mu_n^v$, respectively, and the double bars $\Vert \cdot \Vert$ denotes the Euclidean norm.

To compute the similarity between two scenes, we consider several centroid-based similarity metrics:
\begin{itemize}
\item Relevance-based Quantization closest similarity (\emph{RbQ-c}): the similarity between two scenes $s_u$ and $s_v$ is equal to the largest similarity between their centroids
\begin{equation}
	\max_{m,n} K_{mn}^{uv},
\end{equation}
\item Relevance-based Quantization average similarity  (\emph{RbQ-a}): the similarity between two scenes $s_u$ and $s_v$ is equal to the average of their centroid similarities
\begin{equation}
	\frac{1}{M^2} \sum_{m,n} K_{mn}^{uv}
\end{equation}
and,
\item Relevance-based Quantization weighted similarity  (\emph{RbQ-w}): the similarity between two scenes is computed using a variant of the earth mover's distance applied to the set of centroids each weighted by the number of frames assigned to its cluster.
\end{itemize}

For \emph{RbQ-w}, each scene is represented by a signature
$$
p_u = \lbrace(\mu_1^u,w_1^u),(\mu_2^u,w_2^u),\ldots,(\mu_M^u,w_M^u)\rbrace,
$$
where each of the $M$ centroids $\mu_1^u, \ldots, \mu_M^u$ are paired with corresponding weights $w_1^u, \ldots, w_M^u$. The weight $w_m^u$ for the $m$th centroid $\mu_m^u$ is the number of frames belonging to a particular cluster. The similarity between scenes is then given by a cross-bin histogram distance known as the non-normalized earth mover's distance $\widehat{\EMD}$ introduced by~\cite{pele2008linear}. The $\widehat{\EMD}$ computes the distance between two histograms by finding the minimal cost for transforming one histogram into the other, where cost is measured by the number of transported histogram counts multiplied by a dissimilarity measure between the histogram bins. Here, that measure is given by $1-K_{mn}^{uv}$.

\section{Experiments}
\label{sec:experiments}

To evaluate the representations introduced in the previous section, we apply it to the acoustic scene similarity retrieval task.
Results demonstrate the improved performance of the relevance-based quantization of scattering coefficients compared to baseline methods using summary statistics of MFCCs.
The implementations of the presented methods and the experimental protocol are available online.\footnote{\url{https://github.com/mathieulagrange/paperRelevanceBasedSimilarity}}

\subsection{Dataset}

The experiments in this paper are carried out on the publicly available DCASE 2013 dataset~\cite{7100934}.
Although the dataset was constructed for the task of acoustic scene classification, where the goal is to correctly assign the class of a given recording, we can use the same recordings and class labels for the task of acoustic scene similarity retrieval.
The dataset consists of two parts, a public and a private subset, each made up of $100$ acoustic scene recordings sampled at $44100\,\mathrm{Hz}$ and $30$ seconds in duration. The dataset is evenly divided into $10$ acoustic scene classes: \emph{bus}, \emph{busy street}, \emph{office}, \emph{open air market}, \emph{park}, \emph{quiet street}, \emph{restaurant}, \emph{supermarket}, \emph{tube}, and \emph{tube station}.
The recordings were made by three different recordists at a wide variety of locations in the Greater London area over a period of several months.
In order to avoid any correlation between recording conditions and label distribution, all recordings were carried out under moderate weather conditions, at varying times of day and week, and each recordist recorded each scene type. As a result, the dataset enjoys significant intra-class diversity while remaining of manageable size, making it suitable for evaluation of algorithmic design choices~\cite{lagrange:hal-01082501}.
As an illustration, Figure \ref{fig:scalogram} represents the wavelet scalogram of one recording within the DCASE 2013, labeled as \emph{park}.

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{park_scalogram.png}
\caption{
\label{fig:scalogram}
Wavelet scalogram $\boldsymbol{x}_1(t,\gamma_1)$ of the audio recording \emph{park04} in the DCASE 2013 dataset. We observe that this acoustic scene is a mixture of transient events (chirping birds, footsteps) and stationary texture (flowing water).}
\end{center}
\end{figure}

\subsection{Feature design}

We perform our experiments using both scattering coefficients and MFCCs. For the scattering transform, each $30$-second scene is described by $128$ vectors of dimension $1367$ computed with half-overlapping windows $\boldsymbol{\phi}(t)$ of duration $T=372\,\mathrm{ms}$, for a total of $24\,\mathrm{s}$. Here we discard $3$ seconds from the beginning and end of the scene to avoid boundary artifacts. We also conduct experiments with and without logarithmic compression of the scattering coefficients (see Section~\ref{sec:logcomp}).

MFCCs are computed for windows of $50\,\mathrm{ms}$ and hops of $25\,\mathrm{ms}$ with full frequency range. The standard configuration of $39$ coefficients coupled with an average-energy measure performs best in preliminary tests, so we use this in the following. We average the coefficients using $250\,\mathrm{ms}$ long non-overlapping windows so that each window represents structures of scales close to that of scattering coefficients.

\subsection{Evaluation and algorithm}

The evaluation is performed on the private part of the DCASE 2013 dataset. As a metric, we use the precision at rank $k$ ($p@k$). This number is computed by taking a query item and counting the number of items of the same class within the $k$ closest neighbors, and then averaging over all query items. We determine these neighbors using one of the proposed similarity measures \emph{RbQ-c}, \emph{RbQ-a}, or \emph{RbQ-w}. We compute $p@k$ for $k=\lbrace 1,\ldots,9\rbrace$, since each class only has $10$ items. Note that $p@1$ is equal to the classification accuracy obtained by the nearest-neighbor classifier in a leave-one-out cross-validation setting.

>>>>

\documentclass[smallextended]{svjour3}
%\usepackage{cite}
% *** GRAPHICS RELATED PACKAGES ***
%
%\usepackage[pdftex]{graphicx}
%
\usepackage[cmex10]{amsmath}

@@

%\usepackage[backend=bibtex,style=numeric-comp,sorting=none,firstinits=true]{biblatex}
%%\bibliography{biblio}
%\renewcommand{\baselinestretch}{2}

@@

\begin{document}

\section{Supplementary material}
